---
title: "Використання реляційної бази даних з R"
teaching: 0
exercises: 0
questions:
  - "Як імпортувати дані, що зберігаються в базі даних SQLite, у датафрейм в R?"
  - "Як записати дані з датафрейму у таблицю SQLite?"
  - "Як створити базу даних SQLite з csv файлів"

objectives:
  - "Встановити пакет RSQLite"
  - "Створити з’єднання з базою даних SQLite"
  - "Виконати запити до бази даних"
  - "Створити нову базу даних і наповнити її даними"
  - "Використовувати функції dplyr для доступу та запитів до бази даних SQLite"

keypoints:
  - "Перший ключовий момент."
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("06-")
source("../bin/download_data.R")
```

## Вступ

Поширеною проблемою при роботі з R є те, що всі операції виконуються в оперативній пам’яті. Тут "пам’ять" означає комп’ютерний пристрій під назвою Random Access Memory (RAM), фізично встановлений у вашому комп’ютері. RAM дозволяє читати дані майже за однаковий час, незалежно від того, де вони фізично розташовані в пам’яті. Це різко контрастує зі швидкістю читання даних з CD, DVD або інших носіїв, де швидкість передачі залежала від того, як швидко обертається диск або рухомий механізм зчитування. На жаль, ваш комп’ютер має обмежену кількість RAM, тому обсяг даних, з якими ви можете працювати, обмежений доступною пам’яттю. До цього моменту ми використовували невеликі набори даних, які легко вміщуються в пам’ять комп’ютера. Але що робити, якщо дані занадто великі, щоб комп’ютер міг обробити їх одразу?

In this case, it is helpful to organze the data into a database stored outside of R before creating
a connection to the database itself. This connection will essentially remove the limitation of memory
because SQL queries can be sent directly from R to the database and return to R only the results that you
have identified as being neccessary for your analysis.

Once we have made the connection to the database, much of what we do will look familiar because the code we will be using is very similar to what we saw in the SQL lesson and earlier episodes of this R lesson.

In this lesson, we will be connecting to an SQLite database, which allows us to send strings containing SQL statements directly from R to the database and recieve the results. In addition, we will be connecting to the database in such a way that we can use 'dplyr' functions to operate directly on the database tables.

## Передумови

First, install and load the neccessary packages. You can install the `RSQLite` package with

```{r, eval=FALSE}
install.packages("RSQLite")
```

Load the packages with

```{r}
library(RSQLite)
library(dplyr)
```

Next, create a variable that contains the location of the SQLite database we are going to use. Here, we are assuming that it is in the current working directory.

```{r}
dbfile <- "data/SN7577.sqlite"
```

## Підключення до SQLite бази даних

Connect to the SQLite database specified by `dbfile`, above, using the `dbConnect` function.

```{r}
mydb <- dbConnect(dbDriver("SQLite"), dbfile)
```

Here, `mydb` represents the connection to the database. It will be specified every time we need to access the database.

Now that we have a connection, we can get a list of the tables in the database.

```{r}
dbListTables(mydb)
```

Our objective here is to bring data from the database into R by sending a query to the database and then asking for the results of that query.

```{r}
# Assign the results of a SQL query to an SQLiteResult object
results <- dbSendQuery(mydb, "SELECT * FROM Question1")

# Return results from a custom object to a dataframe
data <- fetch(results)
```

`data` is a standard R dataframe that can be explored and manipulated.

```{r}
# Return column names
names(data)

# Return description of dataframe structure
str(data)

# Return the second column
data[,2]

# Return the value of the second column, fourth row
data[4,2]

# Return the second column where the value of the column 'key' is greater than 7
data[data$key > 7,2]
```

Once you have retrieved the data you should close the connection.

```{r}
dbClearResult(results)
```

In addition to sending simple queries we can send complex one like a join.
You may want to set this up in a concateneted string first for readability.

```{r}
SQL_query <- paste("SELECT q.value,",
                   "count(*) as how_many",
                   "FROM SN7577 s",
                   "JOIN Question1  q",
                   "ON q.key = s.Q1",
                   "GROUP BY  s.Q1")

results <- dbSendQuery(mydb, SQL_query)

data <- fetch(results)

data

dbClearResult(results)
```

:::::::::::::::::::::::::::::::::::::::::  callout

## Завдання

What happens if you send invalid SQL syntax?

:::::::::::::::  solution

## Відповідь

An error message is returned from SQLite.
Notice that R is just the conduit; it cannot check the SQL syntax.

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

We can also create a new database and add tables to it. Let's base this new dataframe on the Question1 table that can be found in our existing database.

```{r}
# First, use a SQL query to extract the Question1 table from the existing database
results = dbSendQuery(mydb, "SELECT * from Question1")

# Then, store it as a dataframe
Q1 <- fetch(results)
```

Now, we can create the new database and add data to it, either from an external file or a local dataframe.

```{r}
dbfile_new = "data/a_newdb.sqlite"
mydb_new = dbConnect(dbDriver("SQLite"), dbfile_new)

dbWriteTable(conn = mydb_new , name = "SN7577", value = "data/SN7577.csv",
             row.names = FALSE, header = TRUE)

dbWriteTable(conn = mydb_new , name = "Q1", value = Q1,
             row.names = FALSE)
```

## Connecting to a Database for `dplyr` Use

When we want to use `dplyr` functions to operate directly on the database tables,
a different connection method is used.

```{r}
mydb_dplyr <- src_sqlite(path="data/SN7577.sqlite")
```

as is the method for running queries. However using the 'tbl' function we still need to provide avalid SQL string. (?)

```{r}
tbl(mydb_dplyr, sql("SELECT count(*) from SN7577"))
```

The real advantage of using `dplyr` is that once we have stored the table as an object
(here, `SN7577_d`), we can use `dplyr` functions instead of SQL statements.

```{r}
# Store the table as an object
SN7577_d <- tbl(mydb_dplyr, sql("SELECT * FROM SN7577"))

# Explore the object
head(SN7577_d, n = 10)
nrow(SN7577_d)

# Apply dplyr functions to the object
SN7577_d %>%
  filter(numage > 60) %>%
  select(sex, age, numage) %>%
  group_by(sex, age) %>%
  summarize(avg_age = mean(numage))
```

Notice that on the `nrow` command we get NA rather than a count of rows. Thisis because `dplyr` doesn't hold the full table even after the 'Select \* ...'

If you need the row count you can use

```{r}
SN7577_d %>%
  tally()
```

:::::::::::::::::::::::::::::::::::::::::  callout

## Exercise

Store the SN7577 table as an object for `dplyr` use.

Write a query using `dplyr` functions that will return the average age (`numage`) by sex for all records where
the response for Q2 is missing (missing values are indicated by a value of -1).

:::::::::::::::  solution

## Solution

```{r}
SN7577_d <- tbl(mydb_dplyr, sql("SELECT * FROM SN7577"))

SN7577_d %>%
  filter(Q2 == -1)   %>%
  group_by(sex)   %>%
  summarize(avg_age = mean(numage))
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

{% include links.md %}


