---
title: Починаючи з даних
teaching: 50
exercises: 30
source: Rmd
---

```{r setup, include=FALSE}
source("data/download_data.R")
```

:::: instructor

Дві основні цілі цих уроків:

- Щоб переконатися, що учням комфортно працювати з кадрами даних,
  може використовувати позначення дужок для вибору фрагментів/стовпців.
- Виставити учнів як фактори. Їхня поведінка не обов'язково інтуїтивна,
  і тому важливо, щоб вони проходили через неї вперше, коли вони піддаються впливу. Зміст уроку має бути достатнім для того, щоб учні
  уникали поширених помилок з ними.

::::::::::::

::::::::::::::::::::::::::::::::::::::: objectives

- Опишіть, що таке двовимірна таблична структура даних.
- Завантажте зовнішні дані з файлу.csv в кадр даних.
- Підсумувати вміст фрейму даних.
- Значення підмножини з векторів.
- Пояснюйте різницю між фактором і рядком.
- Перетворення між рядками та факторами.
- Упорядкування та перейменування факторів.
- Змініть спосіб обробки рядків символів у кадрі даних.
- Вивчіть та змінюйте формати дат.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- Що таке двовимірна таблична структура даних?
- Як я можу прочитати повний файл csv в R?
- Як я можу отримати основну зведену інформацію про мій набір даних?
- Як я можу змінити спосіб обробки R рядків у моєму наборі даних?
- Чому я хотів би, щоб до рядків ставилися по-різному?
- Як дати представлені в R і як я можу змінити формат?

::::::::::::::::::::::::::::::::::::::::::::::::::

## Що таке двовимірні табличні структури даних?

Фрейми даних - це структура даних_de facto_ для табличних даних у `R` і те, що
ми використовуємо для обробки даних, статистики та побудови графіків.

Фрейм даних - це представлення даних у форматі таблиці, де стовпці
є векторами, які мають однакову довжину. Дата фрейми аналогічні, більш звичній електронній таблиці в таких програмах, як Excel, з однією ключовою відмінністю.
Оскільки стовпці є векторами,
кожен стовпець повинен містити один тип даних (наприклад, символи, цілі числа, коефіцієнти). Наприклад, ось малюнок із зображенням кадру даних, що складається з числа, символу та логічного вектора.

! [] (fig/data-frame.svg) {alt='data frame 3 на 3 зі стовпцями, що показують числові, символьні та логічні значення. '}

Фрейми даних можна створювати вручну, але найчастіше вони генеруються функціями
`read_csv () `або `read_table ()`; іншими словами, при імпорті
електронних таблиць з вашого жорсткого диска (або Інтернету). Тепер ми продемонструємо, як
імпортувати табличні дані за допомогою `read_csv () `.

## Презентація даних SAFI

SAFI (Вивчення африканського зрошення під керівництвом фермерів) - це дослідження, яке вивчає сільське господарство
та методи зрошення в Танзанії та Мозамбіку. Дані опитування
були зібрані за допомогою інтерв'ю, проведених у період з листопада 2016 року до
червня 2017 року. Для цього уроку ми будемо використовувати підмножину доступних даних. Для отримання інформації про повний набір даних для навчання, використаний
на інших уроках цього семінару, див.
[опис набору даних] (https://datacarpentry.org/socialsci-workshop/index.html#data).

Ми будемо використовувати підмножину очищеної версії набору даних, яку
було створено шляхом очищення в OpenRefine (`Data/safi_clean.csv`). У цьому наборі даних відсутні дані кодуються як «NULL», кожен рядок містить інформацію
для одного респондента інтерв'ю, а стовпці
представляють:

| назва/стовпця                                                     | Опис                                                                                                                                                                               |
| ----------------------------------------------------------------- | ---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- |
| ключ\_id                                    | Додано, щоб надати унікальний ідентифікатор для кожного спостереження. (Поле InstanceID також робить це, але воно не так зручно у використанні) |
| село                                                              | Назва села                                                                                                                                                                         |
| співбесіда\дата                                                   | Дата співбесіди                                                                                                                                                                    |
| немає\учасників                                                   | Скільки членів у домогосподарстві?                                                                                                                                                 |
| років\життя                                                       | Скільки років ви живете в цьому селі чи сусідньому селі?                                                                                                                           |
| відповідь\_стіна\_тип | Який тип стін має їхній будинок (зі списку)                                                                                                                     |
| кімнати                                                           | Скільки кімнат в основному будинку використовується для сну?                                                                                                                       |
| член\доцент                                                       | Ви член зрошувальної асоціації?                                                                                                                                                    |
| вплив\_конфлікти                            | Ви постраждали від конфліктів з іншими іригаторами в цьому районі?                                                                                                                 |
| кількість\життя                                                   | Кількість тваринництва, що перебуває у власності.                                                                                                                  |
| позиції\власність                                                 | Які з перерахованих предметів належать домогосподарству? (список)                                                                                               |
| немає\їжі                                                         | Скільки страв люди у вашому домогосподарстві зазвичай їдять за день?                                                                                                               |
| місяців\відсутність\їжа                                           | Вкажіть, на які місяці, в останні 12 місяців ви стикалися з ситуацією, коли у вас не було достатньо їжі, щоб прогодувати домогосподарство?                                         |
| Ідентифікатор                                                     | Унікальний ідентифікатор для подання даних форми                                                                                                                                   |

## Імпортування даних

Ви збираєтеся завантажити дані в пам'ять R за допомогою функції `read_csv () `
з пакета **`readr`**, який є частиною **`tidyverse`**; дізнайтеся
більше про колекцію пакетів **`tidyverse`**
[тут] (https://www.tidyverse.org/).
**`readr`** встановлюється як частина **`tidyverse`** встановлення.
Коли ви завантажуєте **`tidyverse`** (`library (tidyverse) `), завантажуються основні пакети
(пакети, які використовуються в більшості аналізів даних), включаючи **`readr`**.

Однак, перш ніж продовжити, це хороша можливість поговорити про конфлікти.
Деякі пакети, які ми завантажуємо, можуть в кінцевому підсумку вводити імена функцій, які вже
використовуються попередньо завантаженими пакетами R. Наприклад, коли ми завантажимо пакет tidyverse
нижче, ми введемо дві суперечливі функції: `filter () `і `lag ()`.
Це відбувається тому, що `filter` і `lag` вже є функціями, які використовуються пакетом stats
(вже попередньо завантажений у R). Тепер, якщо ми, наприклад
, викликаємо функцію `filter () `, R використовуватиме `dplyr: :filter ()` версії
, а не `stats: :filter () `. Це відбувається тому, що, якщо це конфлікт, за замовчуванням
R використовує функцію з останнього завантаженого пакета. Конфліктні функції
можуть викликати у вас певні проблеми в майбутньому, тому важливо, щоб ми
знали про них, щоб ми могли правильно обробляти їх, якщо хочемо.

Для цього нам просто потрібні такі функції з конфліктного пакету:

- `conflicted: :conflict_scout () `: Показує нам будь-які конфліктні функції.
- `conflict_prefer («function», «package_prefered») `: Дозволяє нам вибрати функцію
  за замовчуванням, яку ми хочемо відтепер.

Також важливо знати, що ми можемо в будь-який час просто викликати функцію
безпосередньо з потрібного нам пакету, наприклад `stats: :filter () `.

Навіть при використанні проекту RStudio може бути важко навчитися
вказувати шляхи до розташування файлів. Введіть пакет **тут**! Тут пакет
створює шляхи щодо каталогу верхнього рівня (вашого проекту RStudio). Ці відносні шляхи
працюють \* незалежно від того, де пов'язаний вихідний файл
знаходиться у вашому проєкті, як проєкти аналізу з даними та звітами в різних підкаталогах.  Це важливий контраст з використанням `setwd () `, яке
залежить від того, як ви впорядкуєте файли на комп'ютері.

<figure>
```{r here-figure, echo = FALSE, fig.show = «hold», fig.alt = «Монстри на роздвоєнні дороги, зі знаками, що говорять тут, а не тут. В одному з напрямків, а не тут, що веде до страшних темні ліси з павуками й абсолютними файловими доріжками, поки другий призводить до сонячної, зеленої луки й міста під веселкою й на вільному від абсолютних файлових доріжок. Мистецтво Еллісон Хорст», out.width=» 100% «}
knitr: :include_graphics (» fig/here_horst.png «)
```
<figcaption>
Кредит зображення: <a href="https://github.com/allisonhorst/stats-illustrations">Еллісон Хорст</a>
</figcaption>
</figure>

Перш ніж ми зможемо використовувати функції `read_csv () `і `here ()`, нам потрібно завантажити пакунки
tidyverse і here.

Also, if you recall, the missing data is encoded as "NULL" in the dataset.
We'll tell it to the function, so R will automatically convert all the "NULL"
entries in the dataset into `NA`.

```{r, eval=TRUE, message=FALSE, purl=FALSE}
library(tidyverse)
library(here)

interviews <- read_csv(
  here("data", "SAFI_clean.csv"), 
  na = "NULL")
```

In the above code, we notice the `here()` function takes folder and file names
as inputs (e.g., `"data"`, `"SAFI_clean.csv"`), each enclosed in quotations
(`""`) and separated by a comma. The `here()` will accept as many names as are
necessary to navigate to a particular file
(e.g., `here("analysis", "data", "surveys", "clean", "SAFI_clean.csv)`).

The `here()` function can accept the folder and file names in an alternate format,
using a slash ("/") rather than commas to separate the names. The two methods are
equivalent, so that `here("data", "SAFI_clean.csv")` and
`here("data/SAFI_clean.csv")` produce the same result. (The slash is used on all
operating systems; backslashes are not used.)

If you were to type in the code above, it is likely that the `read.csv()`
function would appear in the automatically populated list of functions. This
function is different from the `read_csv()` function, as it is included in the
"base" packages that come pre-installed with R. Overall, `read.csv()` behaves
similar to `read_csv()`, with a few notable differences. First, `read.csv()`
coerces column names with spaces and/or special characters to different names
(e.g. `interview date` becomes `interview.date`). Second, `read.csv()`
stores data as a `data.frame`, where `read_csv()` stores data as a different kind of data frame called a `tibble`.
We prefer tibbles because they have nice printing properties among other
desirable qualities. Read more about tibbles
[here](https://tibble.tidyverse.org/).

The second statement in the code above creates a data frame but doesn't output
any data because, as you might recall, assignments (`<-`) don't display
anything. (Note, however, that `read_csv` may show informational
text about the data frame that is created.) If we want to check that our data
has been loaded, we can see the contents of the data frame by typing its name:
`interviews` in the console.

```{r, results="hold", purl=FALSE}
interviews
## Try also
## view(interviews)
## head(interviews)
```

:::::::::::::::::::::::::::::::::::::::::  callout

## Note

`read_csv()` assumes that fields are delimited by commas. However, in several
countries, the comma is used as a decimal separator and the semicolon (;) is
used as a field delimiter. If you want to read in this type of files in R,
you can use the `read_csv2` function. It behaves exactly like `read_csv` but
uses different parameters for the decimal and the field separators. If you are
working with another format, they can be both specified by the user. Check out
the help for `read_csv()` by typing `?read_csv` to learn more. There is also
the `read_tsv()` for tab-separated data files, and `read_delim()` allows you
to specify more details about the structure of your file.

::::::::::::::::::::::::::::::::::::::::::::::::::

Note that `read_csv()` actually loads the data as a tibble.
A tibble is an extension of `R` data frames used by the **`tidyverse`**. When
the data is read using `read_csv()`, it is stored in an object of class
`tbl_df`, `tbl`,  and `data.frame`. You can see the class of an object with

```{r, results="hold", purl=FALSE}
class(interviews)
```

As a `tibble`, the type of data included in each column is listed in an
abbreviated fashion below the column names. For instance, here `key_ID` is a
column of floating point numbers (abbreviated `<dbl>` for the word 'double'),
`village` is a column of characters (`<chr>`) and the `interview_date` is a
column in the "date and time" format (`<dttm>`).

## Inspecting data frames

When calling a `tbl_df` object (like `interviews` here), there is already a lot
of information about our data frame being displayed such as the number of rows,
the number of columns, the names of the columns, and as we just saw the class of
data stored in each column. However, there are functions to extract this
information from data frames.  Here is a non-exhaustive list of some of these
functions. Let's try them out!

Size:

- `dim(interviews)` - returns a vector with the number of rows as the first
  element, and the number of columns as the second element (the **dim**ensions of
  the object)
- `nrow(interviews)` - returns the number of rows
- `ncol(interviews)` - returns the number of columns

Content:

- `head(interviews)` - shows the first 6 rows
- `tail(interviews)` - shows the last 6 rows

Names:

- `names(interviews)` - returns the column names (synonym of `colnames()` for
  `data.frame` objects)

Summary:

- `str(interviews)` - structure of the object and information about the class,
  length and content of each column
- `summary(interviews)` - summary statistics for each column
- `glimpse(interviews)` - returns the number of columns and rows of the tibble,
  the names and class of each column, and previews as many values will fit on the
  screen. Unlike the other inspecting functions listed above, `glimpse()` is not a
  "base R" function so you need to have the `dplyr` or `tibble` packages loaded to
  be able to execute it.

Note: most of these functions are "generic." They can be used on other types of
objects besides data frames or tibbles.

## Subsetting data frames

Our `interviews` data frame has rows and columns (it has 2 dimensions).
In practice, we may not need the entire data frame; for instance, we may only
be interested in a subset of the observations (the rows) or a particular set
of variables (the columns). If we want to
access some specific data from it, we need to specify the "coordinates" (i.e., indices) we
want from it. Row numbers come first, followed by column numbers.

:::::::::::::::::::::::::::::::::::::::::  callout

## Tip

Subsetting a `tibble` with `[` always results in a `tibble`.
However, note this is not true in general for data frames, so be careful!
Different ways of specifying these coordinates can lead to results with
different classes. This is covered in the Software Carpentry lesson
[R for Reproducible Scientific Analysis](https://swcarpentry.github.io/r-novice-gapminder/).

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r, purl=FALSE}
## first element in the first column of the tibble
interviews[1, 1]

## first element in the 6th column of the tibble 
interviews[1, 6]

## first column of the tibble (as a vector)
interviews[[1]]

## first column of the tibble
interviews[1]

## first three elements in the 7th column of the tibble
interviews[1:3, 7]

## the 3rd row of the tibble
interviews[3, ]

## equivalent to head_interviews <- head(interviews)
head_interviews <- interviews[1:6, ]
```

`:` is a special function that creates numeric vectors of integers in increasing
or decreasing order, test `1:10` and `10:1` for instance.

You can also exclude certain indices of a data frame using the "`-`" sign:

```{r, purl=FALSE}
interviews[, -1]          # The whole tibble, except the first column

interviews[-c(7:131), ]   # Equivalent to head(interviews)
```

`tibble`s can be subset by calling indices (as shown previously), but also by
calling their column names directly:

```{r, eval=FALSE, purl=FALSE}
interviews["village"]       # Result is a tibble

interviews[, "village"]     # Result is a tibble

interviews[["village"]]     # Result is a vector

interviews$village          # Result is a vector
```

In RStudio, you can use the autocompletion feature to get the full and correct
names of the columns.

:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

1. Create a tibble (`interviews_100`) containing only the data in
   row 100 of the `interviews` dataset.

Now, continue using `interviews` for each of the following activities:

2. Notice how `nrow()` gave you the number of rows in the tibble?

- Use that number to pull out just that last row in the tibble.
- Compare that with what you see as the last row using `tail()` to make
  sure it's meeting expectations.
- Pull out that last row using `nrow()` instead of the row number.
- Create a new tibble (`interviews_last`) from that last row.

3. Using the number of rows in the interviews dataset that you found in
   question 2, extract the row that is in the middle of the dataset. Store
   the content of this middle row in an object named `interviews_middle`.
   (hint: This dataset has an odd number of rows, so finding the middle is a
   bit trickier than dividing n\_rows by 2. Use the median( ) function and what
   you've learned about sequences in R to extract the middle row!

4. Combine `nrow()` with the `-` notation above to reproduce the behavior of
   `head(interviews)`, keeping just the first through 6th rows of the
   interviews dataset.

:::::::::::::::  solution

## Solution

```{r}
## 1.
interviews_100 <- interviews[100, ]
## 2.
# Saving `n_rows` to improve readability and reduce duplication
n_rows <- nrow(interviews)
interviews_last <- interviews[n_rows, ]
## 3.
interviews_middle <- interviews[median(1:n_rows), ]
## 4.
interviews_head <- interviews[-(7:n_rows), ]
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Factors

R has a special data class, called factor, to deal with categorical data that
you may encounter when creating plots or doing statistical analyses. Factors are
very useful and actually contribute to making R particularly well suited to
working with data. So we are going to spend a little time introducing them.

Factors represent categorical data. They are stored as integers associated with
labels and they can be ordered (ordinal) or unordered (nominal). Factors
create a structured  relation between the different levels (values) of a
categorical variable, such as days of the week or responses to a question in
a survey. This can make it easier to see how one element relates to the
other elements in a column. While factors look (and often behave) like
character vectors, they are actually treated as integer vectors by `R`. So
you need to be very careful when treating them as strings.

Once created, factors can only contain a pre-defined set of values, known as
_levels_. By default, R always sorts levels in alphabetical order. For
instance, if you have a factor with 2 levels:

```{r, purl=TRUE}
respondent_floor_type <- factor(c("earth", "cement", "cement", "earth"))
```

R will assign `1` to the level `"cement"` and `2` to the level `"earth"`
(because `c` comes before `e`, even though the first element in this vector is
`"earth"`). You can see this by using the function `levels()` and you can find
the number of levels using `nlevels()`:

```{r, purl=FALSE}
levels(respondent_floor_type)

nlevels(respondent_floor_type)
```

Sometimes, the order of the factors does not matter. Other times you might want
to specify the order because it is meaningful (e.g., "low", "medium", "high").
It may improve your visualization, or it may be required by a particular type of
analysis. Here, one way to reorder our levels in the `respondent_floor_type`
vector would be:

```{r, results=TRUE, purl=FALSE}
respondent_floor_type # current order

respondent_floor_type <- factor(respondent_floor_type, 
                                levels = c("earth", "cement"))

respondent_floor_type # after re-ordering
```

In R's memory, these factors are represented by integers (1, 2), but are more
informative than integers because factors are self describing: `"cement"`,
`"earth"` is more descriptive than `1`, and `2`. Which one is "earth"? You
wouldn't be able to tell just from the integer data. Factors, on the other hand,
have this information built in. It is particularly helpful when there are many
levels. It also makes renaming levels easier. Let's say we made a mistake and
need to recode "cement" to "brick". We can do this using the `fct_recode()`
function from the **`forcats`** package (included in the **`tidyverse`**) which
provides some extra tools to work with factors.

```{r, results=TRUE, purl=FALSE}
levels(respondent_floor_type)

respondent_floor_type <- fct_recode(respondent_floor_type, brick = "cement")

## as an alternative, we could change the "cement" level directly using the
## levels() function, but we have to remember that "cement" is the second level
# levels(respondent_floor_type)[2] <- "brick"

levels(respondent_floor_type)

respondent_floor_type
```

So far, your factor is unordered, like a nominal variable. R does not know the
difference between a nominal and an ordinal variable. You make your factor an
ordered factor by using the `ordered=TRUE` option inside your factor function.
Note how the reported levels changed from the unordered factor above to the
ordered version below. Ordered levels use the less than sign `<` to denote
level ranking.

```{r, results=TRUE, purl=FALSE}
respondent_floor_type_ordered <- factor(respondent_floor_type, 
                                        ordered = TRUE)

respondent_floor_type_ordered # after setting as ordered factor
```

### Converting factors

If you need to convert a factor to a character vector, you use
`as.character(x)`.

```{r, purl=FALSE}
as.character(respondent_floor_type)
```

Converting factors where the levels appear as numbers (such as concentration
levels, or years) to a numeric vector is a little trickier. The `as.numeric()`
function returns the index values of the factor, not its levels, so it will
result in an entirely new (and unwanted in this case) set of numbers.
One method to avoid this is to convert factors to characters, and then to
numbers. Another method is to use the `levels()` function. Compare:

```{r, purl=TRUE}
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990))

as.numeric(year_fct)                     # Wrong! And there is no warning...

as.numeric(as.character(year_fct))       # Works...

as.numeric(levels(year_fct))[year_fct]   # The recommended way.
```

Notice that in the recommended `levels()` approach, three important steps occur:

- We obtain all the factor levels using `levels(year_fct)`
- We convert these levels to numeric values using `as.numeric(levels(year_fct))`
- We then access these numeric values using the underlying integers of the
  vector `year_fct` inside the square brackets

### Renaming factors

When your data is stored as a factor, you can use the `plot()` function to get a
quick glance at the number of observations represented by each factor level.
Let's extract the `memb_assoc` column from our data frame, convert it into a
factor, and use it to look at the number of interview respondents who were or
were not members of an irrigation association:

```{r factor-plot-default-order, fig.alt="Yes/no bar graph showing number of individuals who are members of irrigation association", purl=TRUE}
## create a vector from the data frame column "memb_assoc"
memb_assoc <- interviews$memb_assoc

## convert it into a factor
memb_assoc <- as.factor(memb_assoc)

## let's see what it looks like
memb_assoc

## bar plot of the number of interview respondents who were
## members of irrigation association:
plot(memb_assoc)
```

Looking at the plot compared to the output of the vector, we can see that in
addition to "no"s and "yes"s, there are some respondents for whom the
information about whether they were part of an irrigation association hasn't
been recorded, and encoded as missing data. These respondents do not appear on the plot.
Let's encode them differently so they can be counted and visualized in our plot.

```{r factor-plot-reorder, fig.alt="Bar plot of association membership, showing missing responses.", purl=TRUE}
## Let's recreate the vector from the data frame column "memb_assoc"
memb_assoc <- interviews$memb_assoc

## replace the missing data with "undetermined"
memb_assoc[is.na(memb_assoc)] <- "undetermined"

## convert it into a factor
memb_assoc <- as.factor(memb_assoc)

## let's see what it looks like
memb_assoc

## bar plot of the number of interview respondents who were
## members of irrigation association:
plot(memb_assoc)
```

:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

- Rename the levels of the factor to have the first letter in uppercase:
  "No","Undetermined", and "Yes".

- Now that we have renamed the factor level to "Undetermined", can you
  recreate the barplot such that "Undetermined" is last (after "Yes")?

:::::::::::::::  solution

## Solution

```{r factor-plot-exercise, fig.alt="bar graph showing number of individuals who are members of irrigation association, including undetermined option"}
## Rename levels.
memb_assoc <- fct_recode(memb_assoc, No = "no",
                         Undetermined = "undetermined", Yes = "yes")
## Reorder levels. Note we need to use the new level names.
memb_assoc <- factor(memb_assoc, levels = c("No", "Yes", "Undetermined"))
plot(memb_assoc)
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Formatting Dates

One of the most common issues that new (and experienced!) R users have is
converting date and time information into a variable that is appropriate and
usable during analyses. A best
practice for dealing with date data is to ensure that each component of your
date is available as a separate variable. In our dataset, we have a
column `interview_date` which contains information about the
year, month, and day that the interview was conducted. Let's
convert those dates into three separate columns.

```{r, eval=FALSE, purl=FALSE}
str(interviews)
```

We are going to use the package **`lubridate`**, , which is included in the **`tidyverse`** installation and should be loaded by default.
However, if we deal with older versions of tidyverse (2022 and ealier), we can manually load it by typing `library(lubridate)`.

If necessary, start by loading the required package:

```{r load-package, message=FALSE, purl=FALSE}
library(lubridate)
```

The lubridate function `ymd()` takes a vector representing year, month, and day,
and converts it to a `Date` vector. `Date` is a class of data recognized by R as
being a date and can be manipulated as such. The argument that the function
requires is flexible, but, as a best practice, is a character vector formatted
as "YYYY-MM-DD".

Let's extract our `interview_date` column and inspect the structure:

```{r, purl=FALSE}
dates <- interviews$interview_date
str(dates)
```

When we imported the data in R, `read_csv()` recognized that this column
contained date information. We can now use the `day()`, `month()` and  `year()`
functions to extract this information from the date, and create new columns in
our data frame to store it:

```{r, purl=FALSE}
interviews$day <- day(dates)
interviews$month <- month(dates)
interviews$year <- year(dates)
interviews
```

Notice the three new columns at the end of our data frame.

In our example above, the `interview_date` column was read in correctly as a
`Date` variable but generally that is not the case. Date columns are often read
in as `character` variables and one can use the `as_date()` function to convert
them to the appropriate `Date/POSIXct`format.

Let's say we have a vector of dates in character format:

```{r, purl=FALSE}
char_dates <- c("7/31/2012", "8/9/2014", "4/30/2016")
str(char_dates)
```

We can convert this vector to dates as :

```{r, purl=FALSE}
as_date(char_dates, format = "%m/%d/%Y")
```

Argument `format` tells the function the order to parse the characters and
identify the month, day and year. The format above is the equivalent of
mm/dd/yyyy. A wrong format can lead to parsing errors or incorrect results.

For example, observe what happens when we use a lower case y instead of upper
case Y for the year.

```{r, purl=FALSE}
as_date(char_dates, format = "%m/%d/%y")
```

Here, the `%y` part of the format stands for a two-digit year instead of a
four-digit year, and this leads to parsing errors.

Or in the following example, observe what happens when the month and day
elements of the format are switched.

```{r, purl=FALSE}
as_date(char_dates, format = "%d/%m/%y")
```

Since there is no month numbered 30 or 31, the first and third dates cannot be
parsed.

We can also use functions `ymd()`, `mdy()` or `dmy()` to convert character
variables to date.

```{r, purl=FALSE}
mdy(char_dates)
```

:::::::::::::::::::::::::::::::::::::::: keypoints

- Use read\_csv to read tabular data in R.
- Use factors to represent categorical data in R.

::::::::::::::::::::::::::::::::::::::::::::::::::


