---
title: Маніпулювання даними за допомогою пакету dplyr
teaching: 25
exercises: 15
source: Rmd
---

```{r setup, include=FALSE}
source("data/download_data.R")
```

:::: instructor

- Цей урок є корисним, якщо у вас є графіка, яка демонструє команди dplyr.
  Ви можете змінити [цю презентацію Google Slides] (https://docs.google.com/presentation/d/1A9abypFdFp8urAe9z7GCMjFr4aPeIb8mZAtJA2F7H0w/edit#slide=id.g652714585f_0_114) та використовувати її для своєї роботи.
- Для цього уроку переконайтеся, що учні впевнено користуються оператором pipe (|>).
- Також іноді виникає деяка плутанина щодо того, якими мають бути аргументи `group_by`
  , а також коли використовувати `filter () `та `select ()`.

::::::::::::

::::::::::::::::::::::::::::::::::::::: objectives

- Опишіть призначення пакета R та пакету\*\*`dplyr`\*\*.
- Виділіть певні стовпці в датасеті за допомогою функції **`dplyr`** `select`.
- Виберіть певні рядки в датасеті відповідно до умов фільтрації за допомогою функції **`dplyr`** `filter`.
- З'єднайте результат однієї з **`dplyr`** функції з введенням іншої функції з оператором 'pipe' %>`%`.
- Додайте нові стовпці дата сету які є функціями наявних стовпців з `mutate`.
- Використовуйте концепцію розділення застосування-комбінування для аналізу даних.
- Використовуйте `summarize`, `group_by` і `count`, щоб розділити набір даних на групи спостережень, застосувати зведену статистику для кожної групи, а потім об'єднати результати.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- Як я можу вибрати певні рядки та/або стовпці з набору даних?
- Як я можу об'єднати кілька команд в одну команду?
- Як створити нові стовпці або видалити наявні стовпці з набору даних?

::::::::::::::::::::::::::::::::::::::::::::::::::

**`dplyr`**  — це пакет, який спрощує роботу з табличними даними, використовуючи обмежений набір функцій, що можна поєднувати для отримання та узагальнення інформації з ваших даних.

Як і **`readr`**, **`dplyr`** є частиною набору пакетів tidyverse. Ці пакети були завантажені
в пам'ять R, коли ми раніше викликали `library (tidyverse) `.

:::::::::::::::::::::::::::::::::::::::::  callout

## Примітка

Пакети в tidyverse, а саме **`dplyr`**, **`tidyr`** та **`ggplot2`**
приймають як британські (наприклад, _summarise_), так і американські (наприклад, _summarize_) варіанти написання
різних назв функцій і опцій. На цьому уроці ми використовуємо
американські написання різних функцій; Однак, можна використовувати регіональний варіант для навчання.

::::::::::::::::::::::::::::::::::::::::::::::::::

## Що таке R-пакет?

Пакет **`dplyr`** надає зручні інструменти для виконання найпоширеніших завдань з обробки даних. Його створено для роботи безпосередньо з датафреймами, а багато поширених операцій оптимізовано завдяки тому, що вони написані компільованою мовою (C++) (адже не всі пакети R написані мовою R!).

Існують також пакети для широкого спектра завдань, зокрема для створення графіків (ggplot2, який ми розглянемо пізніше), завантаження даних із бази NCBI або виконання статистичного аналізу вашого набору даних. Багато таких пакунків є
, розміщеними на **C**Omgensive**R** **a**RChive **n**eWork
(CRAN) за допомогою `install.packages` і завантажуються з нього. Ця функція робить пакет доступним для вашої інсталяції R
за допомогою команди `library () `, як ви робили з `tidyverse` раніше.

Щоб легко отримати доступ до документації для пакета в R або RStudio, скористайтеся
`help (package = «package_name») `.

Щоб дізнатися більше про **`dplyr`** після практичного завдання ви можете ознайомитися з цим
[зручне перетворення даних за допомогою cheatsheet **`dplyr`**](https://github.com/rstudio/cheatsheets/blob/main/data-transformation.pdf).

:::::::::::::::::::::::::::::::::::::::::  callout

## Примітка

Пакети `tidyverse` є альтернативи, включаючи
пакет [`data.table`](https://rdatatable.gitlab.io/data.table/). Див.
[порівняння] (https://mgimond.github.io/rug_2019_12/Index.html)
, наприклад, щоб зрозуміти відмінності між використанням `base`, `tidyverse` і
`data.table`.

::::::::::::::::::::::::::::::::::::::::::::::::::

## Навчання **`dplyr`**

Щоб переконатися, що всі використовуватимуть один і той же набір даних для цього уроку, ми знову прочитаємо
набір даних SAFI, який ми завантажили раніше.

```{r, results="hide", purl=FALSE, message=FALSE}

## завантажити бібліотеку tidyverse
library(tidyverse)
library(here)

інтерв'ю <- read_csv (тут («data», "SAFI_clean.csv «), na = «NULL»)

## перевірте дані
interviews

## попередній перегляд даних
# view (interviews)
```

Ми вивчимо деякі з найпоширеніших функцій **`dplyr`**:

- `select ()`: стовпці підмножини
- `filter()`: вкладені рядки в умовах
- `mutate () `: створити нові стовпці, використовуючи інформацію з інших стовпців
- `group_by () `і `()`: створюємо зведену статистику щодо згрупованих даних
- `arrangere () `: сортувати результати
- `count()`: підрахувати окремі значення

## Selecting columns and filtering rows

To select columns of a dataframe, use `select()`. The first argument to this
function is the dataframe (`interviews`), and the subsequent arguments are the
columns to keep, separated by commas. Alternatively, if you are selecting
columns adjacent to each other, you can use a `:` to select a range of columns,
read as "select columns from \_\_\_ to \_\_\_." You may have done something similar in
the past using subsetting. `select()` is essentially doing the same thing as
subsetting, using a package (`dplyr`) instead of R's base functions.

```{r, results="hide", purl=FALSE}
# to select columns throughout the dataframe
select(interviews, village, no_membrs, months_lack_food)
# to do the same thing with subsetting
interviews[c("village","no_membrs","months_lack_food")]
# to select a series of connected columns
select(interviews, village:respondent_wall_type)
```

To choose rows based on specific criteria, we can use the `filter()` function.
The argument after the dataframe is the condition we want our final
dataframe to adhere to (e.g. village name is Chirodzo):

```{r, purl=FALSE}
# filters observations where village name is "Chirodzo"
filter(interviews, village == "Chirodzo")
```

You may also have noticed that the output from these call doesn't run off the
screen anymore. It's one of the advantages of `tbl_df` (also called tibble),
the central data class in the tidyverse, compared to normal dataframes in R.

We can also specify multiple conditions within the `filter()` function. We can
combine conditions using either "and" or "or" statements. In an "and"
statement, an observation (row) must meet **every** criteria to be included
in the resulting dataframe. To form "and" statements within dplyr, we can  pass
our desired conditions as arguments in the `filter()` function, separated by
commas:

```{r, purl=FALSE}

# filters observations with "and" operator (comma)
# output dataframe satisfies ALL specified conditions
filter(interviews, village == "Chirodzo",
                   rooms > 1,
                   no_meals > 2)
```

We can also form "and" statements with the `&` operator instead of commas:

```{r, purl=FALSE}
# filters observations with "&" logical operator
# output dataframe satisfies ALL specified conditions
filter(interviews, village == "Chirodzo" &
                   rooms > 1 &
                   no_meals > 2)
```

In an "or" statement, observations must meet _at least one_ of the specified conditions.
To form "or" statements we use the logical operator for "or," which is the vertical bar (|):

```{r, purl=FALSE}
# filters observations with "|" logical operator
# output dataframe satisfies AT LEAST ONE of the specified conditions
filter(interviews, village == "Chirodzo" | village == "Ruaca")
```

## Pipes

What if you want to select and filter at the same time? There are three
ways to do this: use intermediate steps, nested functions, or pipes.

With intermediate steps, you create a temporary dataframe and use
that as input to the next function, like this:

```{r, purl=FALSE}
interviews2 <- filter(interviews, village == "Chirodzo")
interviews_ch <- select(interviews2, village:respondent_wall_type)
```

This is readable, but can clutter up your workspace with lots of objects that
you have to name individually. With multiple steps, that can be hard to keep
track of.

You can also nest functions (i.e. one function inside of another), like this:

```{r, purl=FALSE}
interviews_ch <- select(filter(interviews, village == "Chirodzo"),
                         village:respondent_wall_type)
```

This is handy, but can be difficult to read if too many functions are nested, as
R evaluates the expression from the inside out (in this case, filtering, then
selecting).

The last option, _pipes_, are a recent addition to R. Pipes let you take the
output of one function and send it directly to the next, which is useful when
you need to do many things to the same dataset. There are two Pipes in R: 1) `%>%` (called magrittr pipe; made available via the **`magrittr`** package, installed automatically with
**`dplyr`**) or 2) `|>` (called native R pipe and it comes preinstalled with R v4.1.0 onwards). Both the pipes are, by and large, function similarly with a few differences (For more information, check: https://www.tidyverse.org/blog/2023/04/base-vs-magrittr-pipe/). The choice of which pipe to be used can be changed in the Global settings in R studio and once that is done, you can type the pipe with:

- <kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>M</kbd> if you have a PC or <kbd>Cmd</kbd> + <kbd>Shift</kbd> + <kbd>M</kbd> if you have a Mac.

```{r, purl=FALSE}
# the following example is run using magrittr pipe but the output will be same with the native pipe
interviews %>%
    filter(village == "Chirodzo") %>%
    select(village:respondent_wall_type)

#interviews |>
#   filter(village == "Chirodzo") |>
#   select(village:respondent_wall_type)
```

In the above code, we use the pipe to send the `interviews` dataset first
through `filter()` to keep rows where `village` is "Chirodzo", then through
`select()` to keep only the columns from `village` to `respondent_wall_type`. Since `%>%`
takes the object on its left and passes it as the first argument to the function
on its right, we don't need to explicitly include the dataframe as an argument
to the `filter()` and `select()` functions any more.

Some may find it helpful to read the pipe like the word "then". For instance,
in the above example, we take the dataframe `interviews`, _then_ we `filter`
for rows with `village == "Chirodzo"`, _then_ we `select` columns `village:respondent_wall_type`.
The **`dplyr`** functions by themselves are somewhat simple,
but by combining them into linear workflows with the pipe, we can accomplish
more complex data wrangling operations.

If we want to create a new object with this smaller version of the data, we
can assign it a new name:

```{r, purl=FALSE}
interviews_ch <- interviews %>%
    filter(village == "Chirodzo") %>%
    select(village:respondent_wall_type)

interviews_ch

```

Note that the final dataframe (`interviews_ch`) is the leftmost part of this
expression.

:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

Using pipes, subset the `interviews` data to include interviews
where respondents were members of an irrigation association
(`memb_assoc`) and retain only the columns `affect_conflicts`,
`liv_count`, and `no_meals`.

:::::::::::::::  solution

## Solution

```{r}
interviews %>%
    filter(memb_assoc == "yes") %>%
    select(affect_conflicts, liv_count, no_meals)
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Mutate

Frequently you'll want to create new columns based on the values in existing
columns, for example to do unit conversions, or to find the ratio of values in
two columns. For this we'll use `mutate()`.

We might be interested in the ratio of number of household members
to rooms used for sleeping (i.e. avg number of people per room):

```{r, purl=FALSE}
interviews %>%
    mutate(people_per_room = no_membrs / rooms)
```

We may be interested in investigating whether being a member of an
irrigation association had any effect on the ratio of household members
to rooms. To look at this relationship, we will first remove
data from our dataset where the respondent didn't answer the
question of whether they were a member of an irrigation association.
These cases are recorded as "NULL" in the dataset.

To remove these cases, we could insert a `filter()` in the chain:

```{r, purl=FALSE}
interviews %>%
    filter(!is.na(memb_assoc)) %>%
    mutate(people_per_room = no_membrs / rooms)
```

The `!` symbol negates the result of the `is.na()` function. Thus, if `is.na()`
returns a value of `TRUE` (because the `memb_assoc` is missing), the `!` symbol
negates this and says we only want values of `FALSE`, where `memb_assoc` **is
not** missing.

:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

Create a new dataframe from the `interviews` data that meets the following
criteria: contains only the `village` column and a new column called
`total_meals` containing a value that is equal to the total number of meals
served in the household per day on average (`no_membrs` times `no_meals`).
Only the rows where `total_meals` is greater than 20 should be shown in the
final dataframe.

**Hint**: think about how the commands should be ordered to produce this data
frame!

:::::::::::::::  solution

## Solution

```{r}
interviews_total_meals <- interviews %>%
    mutate(total_meals = no_membrs * no_meals) %>%
    filter(total_meals > 20) %>%
    select(village, total_meals)
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Split-apply-combine data analysis and the summarize() function

Many data analysis tasks can be approached using the _split-apply-combine_
paradigm: split the data into groups, apply some analysis to each group, and
then combine the results. **`dplyr`** makes this very easy through the use of
the `group_by()` function.

### The `summarize()` function

`group_by()` is often used together with `summarize()`, which collapses each
group into a single-row summary of that group.  `group_by()` takes as arguments
the column names that contain the **categorical** variables for which you want
to calculate the summary statistics. So to compute the average household size by
village:

```{r, purl=FALSE}
interviews %>%
    group_by(village) %>%
    summarize(mean_no_membrs = mean(no_membrs))
```

You can also group by multiple columns:

```{r, purl=FALSE}
interviews %>%
    group_by(village, memb_assoc) %>%
    summarize(mean_no_membrs = mean(no_membrs))
```

Note that the output is a grouped tibble of nine rows by three columns
which is indicated by the by two first lines with the `#`.
To obtain an ungrouped tibble, use the
`ungroup` function:

```{r, purl=FALSE}
interviews %>%
    group_by(village, memb_assoc) %>%
    summarize(mean_no_membrs = mean(no_membrs)) %>%
    ungroup()
```

Notice that the second line with the `#` that previously indicated the grouping has
disappeared and we now only have a 9x3-tibble without grouping.
When grouping both by `village` and `membr_assoc`, we see rows in our table for
respondents who did not specify whether they were a member of an irrigation
association. We can exclude those data from our table using a filter step.

```{r, purl=FALSE}
interviews %>%
    filter(!is.na(memb_assoc)) %>%
    group_by(village, memb_assoc) %>%
    summarize(mean_no_membrs = mean(no_membrs))
```

Once the data are grouped, you can also summarize multiple variables at the same
time (and not necessarily on the same variable). For instance, we could add a
column indicating the minimum household size for each village for each group
(members of an irrigation association vs not):

```{r, purl=FALSE}
interviews %>%
    filter(!is.na(memb_assoc)) %>%
    group_by(village, memb_assoc) %>%
    summarize(mean_no_membrs = mean(no_membrs),
              min_membrs = min(no_membrs))
```

It is sometimes useful to rearrange the result of a query to inspect the values.
For instance, we can sort on `min_membrs` to put the group with the smallest
household first:

```{r, purl=FALSE}
interviews %>%
    filter(!is.na(memb_assoc)) %>%
    group_by(village, memb_assoc) %>%
    summarize(mean_no_membrs = mean(no_membrs),
              min_membrs = min(no_membrs)) %>%
    arrange(min_membrs)
```

To sort in descending order, we need to add the `desc()` function. If we want to
sort the results by decreasing order of minimum household size:

```{r, purl=FALSE}
interviews %>%
    filter(!is.na(memb_assoc)) %>%
    group_by(village, memb_assoc) %>%
    summarize(mean_no_membrs = mean(no_membrs),
              min_membrs = min(no_membrs)) %>%
    arrange(desc(min_membrs))
```

### Counting

When working with data, we often want to know the number of observations found
for each factor or combination of factors. For this task, **`dplyr`** provides
`count()`. For example, if we wanted to count the number of rows of data for
each village, we would do:

```{r, purl=FALSE}
interviews %>%
    count(village)
```

For convenience, `count()` provides the `sort` argument to get results in
decreasing order:

```{r, purl=FALSE}
interviews %>%
    count(village, sort = TRUE)
```

:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

How many households in the survey have an average of
two meals per day? Three meals per day? Are there any other numbers
of meals represented?

:::::::::::::::  solution

## Solution

```{r}
interviews %>%
   count(no_meals)
```

:::::::::::::::::::::::::

Use `group_by()` and `summarize()` to find the mean, min, and max
number of household members for each village. Also add the number of
observations (hint: see `?n`).

:::::::::::::::  solution

## Solution

```{r}
interviews %>%
  group_by(village) %>%
  summarize(
      mean_no_membrs = mean(no_membrs),
      min_no_membrs = min(no_membrs),
      max_no_membrs = max(no_membrs),
      n = n()
  )
```

:::::::::::::::::::::::::

What was the largest household interviewed in each month?

:::::::::::::::  solution

## Solution

```{r}
# if not already included, add month, year, and day columns
library(lubridate) # load lubridate if not already loaded
interviews %>%
    mutate(month = month(interview_date),
           day = day(interview_date),
           year = year(interview_date)) %>%
    group_by(year, month) %>%
    summarize(max_no_membrs = max(no_membrs))
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: keypoints

- Use the `dplyr` package to manipulate dataframes.
- Use `select()` to choose variables from a dataframe.
- Use `filter()` to choose data based on values.
- Use `group_by()` and `summarize()` to work with subsets of data.
- Use `mutate()` to create new variables.

::::::::::::::::::::::::::::::::::::::::::::::::::


