---
title: Візуалізація даних за допомогою ggplot2
teaching: 80
exercises: 35
source: Rmd
---

```{r setup, include=FALSE}
source("data/download_data.R")
```

:::: instructor

- Цей епізод є загальним оглядом ggplot2 і зосереджується на: (1) ознайомленні з системою шарів у ggplot2, (2) використанні аргументу group у функції aes(), (3) базовому налаштуванні графіків.
- Епізод залежить від даних, створених в епізоді "Маніпулювання даними за допомогою tidyr". Якщо ви не дійшли або не пройшли увесь епізод про tidyr, ви можете надати учасникам доступ до даних, завантаживши їх або швидко створивши за допомогою наведеного нижче коду tidyr. Ймовірно, ви захочете скопіювати код
  в Etherpad.
- Якщо ви пропустили епізод про tidyr, ви можете перейти до розділу експорт даних у тому епізоді.

::::::::::::

::::::::::::::::::::::::::::::::::::::: objectives

- Створити точкові, коробкові та стовпчикові діаграми за допомогою ggplot.
- Задати універсальні параметри графіка.
- Пояснити, що таке фасетування та застосувати його в ggplot.
- Змінити естетику наявного графіка ggplot (включно з підписами осей та кольором).
- Побудувати складні та персоналізовані графіки на основі даних у датафреймі.
- Визначити відмінності між візуалізаціями базового R, lattice та ggplot.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- Які компоненти ggplot?
- Які основні відмінності між базовими графіками R, lattice та ggplot?
- Як створити діаграми розсіювання, коробкові та стовпчикові?
- Як змінити естетику (наприклад, колір, прозорість) графіка?
- Як створити кілька графіків одночасно?

::::::::::::::::::::::::::::::::::::::::::::::::::

Почнемо із завантаження необхідного пакета. **`ggplot2`** також входить до складу пакету **`tidyverse`**.

```{r load-package, message=FALSE, purl=FALSE}
library(tidyverse)
```

Якщо дані ще не знаходяться в робочій області, завантажте ті, які ми зберегли в попередньому уроці.

```{r load-data, purl=FALSE}
interviews_plotting <- read_csv("data_output/interviews_plotting.csv")
```

Якщо ви не змогли завершити попередній урок або не зберегли дані, ви можете створити їх зараз. Або завантажте за допомогою `read_csv()` (варіант 1), або створіть за допомогою **dplyr** та **tidyr** (варіант 2).

::: tab

### Варіант 1: Завантажте дані

```{r, purl=FALSE, eval=FALSE}
interviews_plotting <- read_csv("https://raw.githubusercontent.com/datacarpentry/r-socialsci/main/episodes/data/interviews_plotting.csv")
```

### Варіант 2: Створіть дані

```{r, purl=FALSE, eval=FALSE}
## Може бути використано для завантаження даних з попереднього уроку!
interviews_plotting <- interviews %>%
  ## pivot wider by items_owned
  separate_longer_delim(items_owned, delim = ";") %>%
  replace_na(list(items_owned = "no_listed_items")) %>%
  ## Use of grouped mutate to find number of rows
  group_by(key_ID) %>% 
  mutate(items_owned_logical = TRUE,
         number_items = if_else(items_owned == "no_listed_items", 0, n())) %>% 
  pivot_wider(names_from = items_owned,
              values_from = items_owned_logical,
              values_fill = list(items_owned_logical = FALSE)) %>% 
  ## pivot wider by months_lack_food
  separate_longer_delim(months_lack_food, delim = ";") %>%
  mutate(months_lack_food_logical = TRUE,
         number_months_lack_food = if_else(months_lack_food == "none", 0, n())) %>%
  pivot_wider(names_from = months_lack_food,
              values_from = months_lack_food_logical,
              values_fill = list(months_lack_food_logical = FALSE))
```

:::

## Параметри візуалізації в R

Перед тим як почати з **`ggplot2`**, корисно знати, що в R існує кілька способів створювати візуалізації. Хоча **`ggplot2`** чудово підходить для створення складних та висококастомізованих графіків, існують простіші та швидші альтернативи, з якими ви можете зіткнутися або використовувати залежно від контексту. Давайте коротко розглянемо декілька з них:

### Базові графіки R

Базові графіки R є найпростішою формою візуалізації й чудово підходять для швидкого, дослідницького аналізу. Можна створювати графіки за допомогою невеликої кількості коду, але їхнє налаштування може бути громіздким у порівнянні з **`ggplot2`**.

Приклад простого точкового графіка у базовому R з використанням змінних `no_membrs` та `liv_count`:

```{r, purl=FALSE}
plot(interviews_plotting$no_membrs, interviews_plotting$liv_count,
     main = "Base R Scatterplot",
     xlab = "Number of Household Members",
     ylab = "Number of Livestock Owned")
```

### **`Lattice`**

Lattice — це ще одна система побудови графіків в R, яка дозволяє легко створювати багато панельні графіки у вигляді решітки. Воно відрізняється від ggplot2 тим, що весь графік визначається одним викликом функції та можливості для змін після побудови графіка обмежені.

Приклад графіка **lattice** з використанням змінних `no_membrs` та `liv_count`, розбитих за `village`:

```{r lattice, message=FALSE, purl=FALSE}
library(lattice)
```

```{r, purl=FALSE}
xyplot(liv_count ~ no_membrs | village, data = interviews_plotting,
       main = "Lattice Plot: Livestock Count by Household Members",
       xlab = "Number of Household Members",
       ylab = "Number of Livestock Owned")
```

## Побудова графіків за допомогою **`ggplot2`**

**`ggplot2`** — це пакет для побудови графіків у R, який дозволяє легко створювати складні візуалізації з даних датафрейму. Він забезпечує програмний інтерфейс, який дозволяє задавати, які змінні зображати, як їх показувати та загальні візуальні властивості. Тому нам потрібно внести лише мінімальні зміни, якщо вихідні дані зміняться або якщо ми вирішимо перейти, наприклад, від стовпчикової діаграми до точкової (діаграма розсіювання). Це допомагає створювати графіки публікаційної якості з мінімальною кількістю коригувань і налаштувань.

Функції **`ggplot2`** найкраще працюють із даними у 'long' форматі, тобто коли кожен вимір представлений окремим стовпцем, а кожне спостереження — окремим рядком. Добре структуровані дані заощадять вам багато часу під час побудови графіків у **`ggplot2`**

ggplot графіки будуються крок за кроком шляхом додавання нових елементів. Такий спосіб додавання шарів забезпечує велику гнучкість і дозволяє налаштовувати побудовані графіки.

Кожен графік, побудований за допомогою ggplot2, повинен включати такі елементи

- Дані

- Естетичне відображення (aes)

  - Описує як змінні накладаються на графічні атрибути
  - До таких атрибутів належать осі x і y, колір, заливка, форма точок, прозорість (alpha)

- Геометричні об'єкти (geom)

  - Визначає, як дані будуть зображені графічно — у вигляді стовпчиків (`geom_bar`), діаграми розсіювання (`geom_point`), лінії (`geom_line`) тощо.

Таким чином, шаблон побудови графіка в ggplot2:

```
<DATA> %>%
    ggplot(aes(<MAPPINGS>)) +
    <GEOM_FUNCTION>()
```

Нагадаємо з попереднього уроку, що оператор `%>%` (pipe) передає результат попереднього кроку як вхідні дані для наступної функції. **`ggplot`** — це функція, яка очікує датафрейм як перший аргумент. Це дозволяє не вказувати аргумент `data =` всередині функції `ggplot`, а просто передавати дані через pipe (`%>%`).

- використайте функцію `ggplot()` та прив'яжіть графік до певного датафрейму.

```{r ggplot-steps-1, eval=FALSE, purl=FALSE}
interviews_plotting %>%
    ggplot()
```

- визначити відображення (mapping) за допомогою функції `aes()`, обравши змінні для побудови графіка та вказавши, як їх показувати на графіку, наприклад, через позиції по осях x/y або через характеристики, такі як розмір, форма, колір тощо.

```{r ggplot-steps-2, eval=FALSE, purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = no_membrs, y = number_items))
```

- додати 'geoms' — графічні представлення даних на графіку (точки, лінії, стовпчики). **`ggplot2`** пропонує багато різних geoms; сьогодні ми використаємо деякі з поширених, зокрема:

  - `geom_point()` для діаграм розсіювання, точкових графіків тощо.
  - `geom_boxplot()` для коробкових діаграм (boxplots)!
  - `geom_line()` для ліній тренду, часових рядів тощо.

Щоб додати geom до графіка, використовуйте оператор `+`. Оскільки у нас дві безперервні змінні, спочатку використаємо `geom_point()`:

```{r first-ggplot, purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = no_membrs, y = number_items)) +
    geom_point()
```

Символ `+` у пакеті **`ggplot2`** особливо корисний, тому що він дозволяє змінювати вже створені об’єкти ggplot. Це означає, що ви можете легко налаштовувати шаблони графіків і зручно досліджувати різні типи візуалізацій, тому наведений вище графік також можна створити за допомогою такого коду, подібного до підходу "проміжних кроків" у попередньому уроці:

```{r first-ggplot-with-plus, fig.alt="Scatter plot of number of items owned versus number of household members.", eval=FALSE, purl=FALSE}
# Присвоїти графік змінній
interviews_plot <- interviews_plotting %>%
    ggplot(aes(x = no_membrs, y = number_items))

# Побудувати точковий графік
interviews_plot +
    geom_point()
```

:::::::::::::::::::::::::::::::::::::::::  callout

## Notes

- Anything you put in the `ggplot()` function can be seen by any geom layers
  that you add (i.e., these are universal plot settings). This includes the x-
  and y-axis mapping you set up in `aes()`.
- You can also specify mappings for a given geom independently of the mapping
  defined globally in the `ggplot()` function.
- The `+` sign used to add new layers must be placed at the end of the line
  containing the _previous_ layer. If, instead, the `+` sign is added at the
  beginning of the line containing the new layer, **`ggplot2`** will not add
  the new layer and will return an error message.

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r ggplot-with-plus-position, eval=FALSE, purl=FALSE}
## This is the correct syntax for adding layers
interviews_plot +
    geom_point()

## This will not add the new layer and will return an error message
interviews_plot
+ geom_point()
```

## Building your plots iteratively

Building plots with **`ggplot2`** is typically an iterative process. We start by
defining the dataset we'll use, lay out the axes, and choose a geom:

```{r create-ggplot-object, fig.alt="Scatter plot of number of items owned versus number of household members.", purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = no_membrs, y = number_items)) +
    geom_point()
```

Then, we start modifying this plot to extract more information from it. For
instance, when inspecting the plot we notice that points only appear at the
intersection of whole numbers of `no_membrs` and `number_items`. Also, from a
rough estimate, it looks like there are far fewer dots on the plot than there
rows in our dataframe. This should lead us to believe that there may be multiple
observations plotted on top of each other (e.g. three observations where
`no_membrs` is 3 and `number_items` is 1).

There are two main ways to alleviate overplotting issues:

1. changing the transparency of the points
2. jittering the location of the points

Let's first explore option 1, changing the transparency of the points. What we
mean when we say "transparency" we mean the opacity of point, or your ability to
see through the point. We can control the transparency of the points with the
`alpha` argument to `geom_point`. Values of `alpha` range from 0 to 1, with
lower values corresponding to more transparent colors (an `alpha` of 1 is the
default value). Specifically, an alpha of 0.1, would make a point one-tenth as
opaque as a normal point. Stated differently ten points stacked on top of
each other would correspond to a normal point.

Here, we change the `alpha` to 0.5, in an attempt to help fix the overplotting.
While the overplotting isn't solved, adding transparency begins to address this
problem, as the points where there are overlapping observations are darker (as
opposed to lighter gray):

```{r adding-transparency, fig.alt="Scatter plot of number of items owned versus number of household members, with transparency added to points.", purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = no_membrs, y = number_items)) +
    geom_point(alpha = 0.5)
```

That only helped a little bit with the overplotting problem, so let's try option
two. We can jitter the points on the plot, so that we can see each point in the
locations where there are overlapping points. Jittering introduces a little bit
of randomness into the position of our points. You can think of this process as
taking the overplotted graph and giving it a tiny shake. The points will move a
little bit side-to-side and up-and-down, but their position from the original
plot won't dramatically change. Note that this solution is suitable for plotting
integer figures, while for numeric figures with decimals, geom_jitter() becomes
inappropriate because it obscures the true value of the observation.

We can jitter our points using the `geom_jitter()` function instead of the
`geom_point()`  function, as seen below:

```{r adding-jitter, fig.alt="Scatter plot of number of items owned versus number of household members, showing jitter.", purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = no_membrs, y = number_items)) +
    geom_jitter()
```

The `geom_jitter()` function allows for us to specify the amount of random
motion in the jitter, using the `width` and `height` arguments. When we don't
specify values for `width` and `height`, `geom_jitter()` defaults to 40% of the
resolution of the data (the smallest change that can be measured). Hence, if we
would like _less_ spread in our jitter than was default, we should pick values
between 0.1 and 0.4. Experiment with the values to see how your plot changes.

```{r adding-width-height, fig.alt="Scatter plot of number of items owned versus number of household members, with jitter and transparency.", purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = no_membrs, y = number_items)) +
    geom_jitter(alpha = 0.5,
                width = 0.2,
                height = 0.2)
```

For our final change, we can also add colours for all the points by specifying
a `color` argument inside the `geom_jitter()` function:

```{r adding-colors, fig.alt="Scatter plot of number of items owned versus number of household members, showing points as blue.", purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = no_membrs, y = number_items)) +
    geom_jitter(alpha = 0.5,
                color = "blue",
                width = 0.2,
                height = 0.2)
```

To colour each village in the plot differently, you could use a vector as an input
to the argument **`color`**.  However, because we are now mapping features of the
data to a colour, instead of setting one colour for all points, the colour of the
points now needs to be set inside a call to the **`aes`** function. When we map
a variable in our data to the colour of the points, **`ggplot2`** will provide a
different colour corresponding to the different values of the variable. We will
continue to specify the value of **`alpha`**, **`width`**, and **`height`**
outside of the **`aes`** function because we are using the same value for
every point. ggplot2 understands both the Commonwealth English and
American English spellings for colour, i.e., you can use either `color`
or `colour`. Here is an example where we color points by the **`village`**
of the observation:

```{r color-by-species, purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = no_membrs, y = number_items)) +
    geom_jitter(aes(color = village), alpha = 0.5, width = 0.2, height = 0.2)
```

There appears to be a positive trend between number of household
members and number of items owned (from the list provided). Additionally,
this trend does not appear to be different by village.

:::::::::::::::::::::::::::::::::::::::::  callout

## Notes

As you will learn, there are multiple ways to plot the a relationship
between variables. Another way to plot data with overlapping points is
to use the `geom_count` plotting function. The `geom_count()`  function
makes the size of each point representative of the number of data items
of that type and the legend gives point sizes associated to particular
numbers of items.

```{r color-by-species-notes, fig.alt="Previous plot with dots colored by village.", purl=FALSE}
interviews_plotting %>%
   ggplot(aes(x = no_membrs, y = number_items, color = village)) +
   geom_count()
```

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

Use what you just learned to create a scatter plot of `rooms` by `village`
with the `respondent_wall_type` showing in different colours. Does this
seem like a good way to display the relationship between these variables?
What other kinds of plots might you use to show this type of data?

:::::::::::::::  solution

## Solution

```{r scatter-challenge, fig.alt="Scatter plot showing positive trend between number of household members and number of items owned.", answer=TRUE, purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = village, y = rooms)) +
    geom_jitter(aes(color = respondent_wall_type),
	    alpha = 0.5,
		    width = 0.2,
		    height = 0.2)
```

This is not a great way to show this type of data because it is difficult to
distinguish between villages. What other plot types could help you visualize
this relationship better?

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Boxplot

We can use boxplots to visualize the distribution of rooms for each
wall type:

```{r boxplot, fig.alt="Box plot of number of rooms by wall type.", purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = respondent_wall_type, y = rooms)) +
    geom_boxplot()
```

By adding points to a boxplot, we can have a better idea of the number of
measurements and of their distribution:

```{r boxplot-with-jitter, fig.alt="Previous plot with dot plot added as additional layer to show individual values. Boxplot layer is transparent.", purl=FALSE}
interviews_plotting %>%
    ggplot(aes(x = respondent_wall_type, y = rooms)) +
    geom_boxplot(alpha = 0) +
    geom_jitter(alpha = 0.5,
    		color = "tomato",
    		width = 0.2,
    		height = 0.2)
```

We can see that muddaub houses and sunbrick houses tend to be smaller than
burntbrick houses.

Notice how the  boxplot layer is behind the jitter layer? What do you need to
change in the code to put the boxplot layer in front of the jitter layer?

:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

Boxplots are useful summaries, but hide the _shape_ of the distribution. For
example, if the distribution is bimodal, we would not see it in a
boxplot. An alternative to the boxplot is the violin plot, where the shape
(of the density of points) is drawn.

- Replace the box plot with a violin plot; see `geom_violin()`.

:::::::::::::::  solution

## Solution

```{r violin-plot}
interviews_plotting %>%
  ggplot(aes(x = respondent_wall_type, y = rooms)) +
  geom_violin(alpha = 0) +
  geom_jitter(alpha = 0.5, color = "tomato")
```

:::::::::::::::::::::::::

So far, we've looked at the distribution of room number within wall type. Try
making a new plot to explore the distribution of another variable within wall
type.

- Create a boxplot for `liv_count` for each wall type. Overlay the boxplot
  layer on a jitter layer to show actual measurements.

:::::::::::::::  solution

## Solution

```{r boxplot-exercise, fig.alt="Box plot of number of livestock owned by wall type, with dot plot added as additional layer to show individual values."}
interviews_plotting %>%
   ggplot(aes(x = respondent_wall_type, y = liv_count)) +
   geom_boxplot(alpha = 0) +
   geom_jitter(alpha = 0.5, width = 0.2, height = 0.2)
```

:::::::::::::::::::::::::

- Add colour to the data points on your boxplot according to whether the
  respondent is a member of an irrigation association (`memb_assoc`).

:::::::::::::::  solution

## Solution

```{r boxplot-exercise-factor, fig.alt="Previous plot with dots colored based on whether respondent was a member of an irrigation association."}
interviews_plotting %>%
  ggplot(aes(x = respondent_wall_type, y = liv_count)) +
  geom_boxplot(alpha = 0) +
  geom_jitter(aes(color = memb_assoc), alpha = 0.5, width = 0.2, height = 0.2)
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Barplots

Barplots are also useful for visualizing categorical data. By default,
`geom_bar` accepts a variable for x, and plots the number of instances each
value of x (in this case, wall type) appears in the dataset.

```{r barplot-1, fig.alt="Bar plot showing counts of respondent wall types."}
interviews_plotting %>%
    ggplot(aes(x = respondent_wall_type)) +
    geom_bar()
```

We can use the `fill` aesthetic for the `geom_bar()` geom to colour bars by
the portion of each count that is from each village.

```{r barplot-stack, fig.alt="Stacked bar plot of wall types showing each village as a different color."}
interviews_plotting %>%
    ggplot(aes(x = respondent_wall_type)) +
    geom_bar(aes(fill = village))
```

This creates a stacked bar chart. These are generally more difficult to read
than side-by-side bars. We can separate the portions of the stacked bar that
correspond to each village and put them side-by-side by using the `position`
argument for `geom_bar()` and setting it to "dodge".

```{r barplot-dodge, fig.alt="Bar plot of respondent wall types with each village as a separate bar."}
interviews_plotting %>%
    ggplot(aes(x = respondent_wall_type)) +
    geom_bar(aes(fill = village), position = "dodge")
```

This is a nicer graphic, but we're more likely to be interested in the
proportion of each housing type in each village than in the actual count of
number of houses of each type (because we might have sampled different numbers
of households in each village). To compare proportions, we will first create a
new data frame (`percent_wall_type`) with a new column named "percent"
representing the percent of each house type in each village. We will remove
houses with cement walls, as there was only one in the dataset.

```{r wall-type-data}
percent_wall_type <- interviews_plotting %>%
    filter(respondent_wall_type != "cement") %>%
    count(village, respondent_wall_type) %>%
    group_by(village) %>%
    mutate(percent = (n / sum(n)) * 100) %>%
    ungroup()
```

Now we can use this new data frame to create our plot showing the
percentage of each house type in each village.

```{r barplot-wall-type, fig.alt="Side by side bar plot showing percent of respondents in each village with each wall type."}
percent_wall_type %>%
    ggplot(aes(x = village, y = percent, fill = respondent_wall_type)) +
    geom_bar(stat = "identity", position = "dodge")
```

:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

Create a bar plot showing the proportion of respondents in each
village who are or are not part of an irrigation association
(`memb_assoc`). Include only respondents who answered that question
in the calculations and plot. Which village had the lowest proportion of
respondents in an irrigation association?

:::::::::::::::  solution

## Solution

```{r barplot-memb-assoc, fig.alt="Bar plot showing percent of respondents in each village who were part of association."}
percent_memb_assoc <- interviews_plotting %>%
  filter(!is.na(memb_assoc)) %>%
  count(village, memb_assoc) %>%
  group_by(village) %>%
  mutate(percent = (n / sum(n)) * 100) %>%
  ungroup()

percent_memb_assoc %>%
   ggplot(aes(x = village, y = percent, fill = memb_assoc)) +
    geom_bar(stat = "identity", position = "dodge")
```

Ruaca had the lowest proportion of members in an irrigation association.

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

## Adding Labels and Titles

By default, the axes labels on a plot are determined by the name of the variable
being plotted. However, **`ggplot2`** offers lots of customization options,
like specifying the axes labels, and adding a title to the plot with
relatively few lines of code. We will add more informative x-and y-axis
labels to our plot, a more explanatory label to the legend, and a plot title.

The `labs` function takes the following arguments:

- `title` -- to produce a plot title
- `subtitle` -- to produce a plot subtitle (smaller text placed beneath the title)
- `caption` -- a caption for the plot
- `...` -- any pair of name and value for aesthetics used in the plot (e.g.,
  `x`, `y`, `fill`, `color`, `size`)

```{r barplot-wall-types-labeled, fig.alt="Previous plot with plot title and labells added."}
percent_wall_type %>%
    ggplot(aes(x = village, y = percent, fill = respondent_wall_type)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title = "Proportion of wall type by village",
         fill = "Type of Wall in Home",
         x = "Village",
         y = "Percent")
```

## Faceting

Rather than creating a single plot with side-by-side bars for each
village, we may want to create multiple plot, where each plot shows the
data for a single village. This would be especially useful if we had
a large number of villages that we had sampled, as a large number of
side-by-side bars will become more difficult to read.

**`ggplot2`** has a special technique called _faceting_ that allows the
user to split one plot into multiple plots based on a factor included
in the dataset. We will use it to split our barplot of housing type
proportion by village so that each village has its own panel in a
multi-panel plot:

```{r barplot-faceting, fig.alt="Bar plot showing percent of each wall type in each village."}
percent_wall_type %>%
    ggplot(aes(x = respondent_wall_type, y = percent)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title="Proportion of wall type by village",
         x="Wall Type",
         y="Percent") +
    facet_wrap(~ village)
```

Click the "Zoom" button in your RStudio plots pane to view a larger
version of this plot.

Usually plots with white background look more readable when printed.  We can set
the background to white using the function `theme_bw()`. Additionally, you can remove
the grid:

```{r barplot-theme-bw, fig.alt="Bar plot showing percent of each wall type in each village, with black and white theme applied.", purl=FALSE}
percent_wall_type %>%
    ggplot(aes(x = respondent_wall_type, y = percent)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title="Proportion of wall type by village",
         x="Wall Type",
         y="Percent") +
    facet_wrap(~ village) +
    theme_bw() +
    theme(panel.grid = element_blank())
```

What if we wanted to see the proportion of respondents in each village
who owned a particular item? We can calculate the percent of people
in each village who own each item and then create a faceted series of
bar plots where each plot is a particular item. First we need to
calculate the percentage of people in each village who own each item:

```{r percent-items-data}
percent_items <- interviews_plotting %>%
    group_by(village) %>%
    summarize(across(bicycle:no_listed_items, ~ sum(.x) / n() * 100)) %>%
    pivot_longer(bicycle:no_listed_items, names_to = "items", values_to = "percent")
```

To calculate this percentage data frame, we needed to use the `across()`
function within a `summarize()` operation. Unlike the previous example with a
single wall type variable, where each response was exactly one of the types
specified, people can (and do) own more than one item. So there are multiple
columns of data (one for each item), and the percentage calculation needs to be
repeated for each column.

Combining `summarize()` with `across()` allows us to specify first, the columns
to be summarized (`bicycle:no_listed_items`) and then the calculation. Because
our calculation is a bit more complex than is available in a built-in function,
we define a new formula:

- `~` indicates that we are defining a formula,
- `sum(.x)` gives the number of people owning that item by counting the number of `TRUE`
  values (`.x` is shorthand for the column being operated on),
- and `n()` gives the current group size.

After the `summarize()` operation, we have a table of percentages with each item
in its own column, so a `pivot_longer()` is required to transform the table into
an easier format for plotting. Using this data frame, we can now create a
multi-paneled bar plot.

```{r percent-items-barplot, fig.alt="Multi-panel bar chart showing percent  of respondents in each village and who owned each item, with no grids behid bars."}
percent_items %>%
    ggplot(aes(x = village, y = percent)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~ items) +
    theme_bw() +
    theme(panel.grid = element_blank())
```

## **`ggplot2`** themes

In addition to `theme_bw()`, which changes the plot background to white,
**`ggplot2`** comes with several other themes which can be useful to quickly
change the look of your visualization. The complete list of themes is available
at [https://ggplot2.tidyverse.org/reference/ggtheme.html](https://ggplot2.tidyverse.org/reference/ggtheme.html). `theme_minimal()` and
`theme_light()` are popular, and `theme_void()` can be useful as a starting
point to create a new hand-crafted theme.

The
[ggthemes](https://jrnold.github.io/ggthemes/reference/index.html)
package provides a wide variety of options (including an Excel 2003 theme). The
[**`ggplot2`** extensions website](https://exts.ggplot2.tidyverse.org/) provides a list
of packages that extend the capabilities of **`ggplot2`**, including additional
themes.

:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

Experiment with at least two different themes. Build the previous plot
using each of those themes. Which do you like best?

::::::::::::::::::::::::::::::::::::::::::::::::::

## Customization

Take a look at the ,
and think of ways you could improve the plot.

Now, let's change names of axes to something more informative than 'village' and
'percent' and add a title to the figure:

```{r ggplot-customization, purl=FALSE}
percent_items %>%
    ggplot(aes(x = village, y = percent)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~ items) +
    labs(title = "Percent of respondents in each village who owned each item",
         x = "Village",
         y = "Percent of Respondents") +
    theme_bw()
```

The axes have more informative names, but their readability can be improved by
increasing the font size:

```{r ggplot-customization-font-size, purl=FALSE}
percent_items %>%
    ggplot(aes(x = village, y = percent)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~ items) +
    labs(title = "Percent of respondents in each village who owned each item",
         x = "Village",
         y = "Percent of Respondents") +
    theme_bw() +
    theme(text = element_text(size = 16))
```

Note that it is also possible to change the fonts of your plots. If you are on
Windows, you may have to install the , and follow the instructions included
in the README for this package.

After our manipulations, you may notice that the values on the x-axis are still
not properly readable. Let's change the orientation of the labels and adjust
them vertically and horizontally so they don't overlap. You can use a 90-degree
angle, or experiment to find the appropriate angle for diagonally oriented
labels. With a larger font, the title also runs off. We can add "\\n" in the string
for the title to insert a new line:

```{r ggplot-customization-label-orientation, fig.alt="Multi-panel bar charts showing percent of respondents in each village and who owned each item, with grids behind the bars.", purl=FALSE}
percent_items %>%
    ggplot(aes(x = village, y = percent)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~ items) +
    labs(title = "Percent of respondents in each village \n who owned each item",
         x = "Village",
         y = "Percent of Respondents") +
    theme_bw() +
    theme(axis.text.x = element_text(colour = "grey20", size = 12, angle = 45,
                                     hjust = 0.5, vjust = 0.5),
          axis.text.y = element_text(colour = "grey20", size = 12),
          text = element_text(size = 16))
```

If you like the changes you created better than the default theme, you can save
them as an object to be able to easily apply them to other plots you may create.
We can also add `plot.title = element_text(hjust = 0.5)` to centre the title:

```{r ggplot-custom-themes, purl=FALSE}
grey_theme <- theme(axis.text.x = element_text(colour = "grey20", size = 12,
                                               angle = 45, hjust = 0.5,
                                               vjust = 0.5),
                    axis.text.y = element_text(colour = "grey20", size = 12),
                    text = element_text(size = 16),
                    plot.title = element_text(hjust = 0.5))


percent_items %>%
    ggplot(aes(x = village, y = percent)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~ items) +
    labs(title = "Percent of respondents in each village \n who owned each item",
         x = "Village",
         y = "Percent of Respondents") +
    grey_theme
```

:::::::::::::::::::::::::::::::::::::::  challenge

## Exercise

With all of this information in hand, please take another five minutes to
either improve one of the plots generated in this exercise or create a
beautiful graph of your own. Use the RStudio [**`ggplot2`** cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/data-visualization-2.1.pdf)
for inspiration. Here are some ideas:

- See if you can make the bars white with black outline.
- Try using a different colour palette (see
  [http://www.cookbook-r.com/Graphs/Colors\_(ggplot2)/](https://www.cookbook-r.com/Graphs/Colors_\\(ggplot2\\)/)).

::::::::::::::::::::::::::::::::::::::::::::::::::

After creating your plot, you can save it to a file in your favourite format. The Export tab in the **Plot** pane in RStudio will save your plots at low resolution, which will not be accepted by many journals and will not scale well for posters.

Instead, use the `ggsave()` function, which allows you to easily change the dimension and resolution of your plot by adjusting the appropriate arguments (`width`, `height` and `dpi`).

Make sure you have the `fig_output/` folder in your working directory.

```{r ggsave-example, eval=FALSE, purl=FALSE}
my_plot <- percent_items %>%
    ggplot(aes(x = village, y = percent)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~ items) +
    labs(title = "Percent of respondents in each village \n who owned each item",
         x = "Village",
         y = "Percent of Respondents") +
    theme_bw() +
    theme(axis.text.x = element_text(color = "grey20", size = 12, angle = 45,
                                     hjust = 0.5, vjust = 0.5),
          axis.text.y = element_text(color = "grey20", size = 12),
          text = element_text(size = 16),
          plot.title = element_text(hjust = 0.5))

ggsave("fig_output/name_of_file.png", my_plot, width = 15, height = 10)
```

Note: The parameters `width` and `height` also determine the font size in the saved plot.

:::::::::::::::::::::::::::::::::::::::: keypoints

- `ggplot2` is a flexible and useful tool for creating plots in R.
- The data set and coordinate system can be defined using the `ggplot` function.
- Additional layers, including geoms, are added using the `+` operator.
- Boxplots are useful for visualizing the distribution of a continuous variable.
- Barplots are useful for visualizing categorical data.
- Faceting allows you to generate multiple plots based on a categorical variable.

::::::::::::::::::::::::::::::::::::::::::::::::::


