---
title: Обробка даних JSON (необов'язково)
teaching: 30
exercises: 15
source: Rmd
---

```{r setup, include=FALSE}
source("data/download_data.R")
library(tidyverse)
```

:::: instructor

- Цей урок необов’язковий і призначений для ознайомлення з даними у форматі JSON, зокрема з тим, як зчитувати JSON у R та конвертувати ці дані у датафрейм або масив.
- Зауважте, що урок був створений спільнотою і досі знаходиться у процесі доопрацювання. Тому він може покращитися завдяки відгукам викладачів та/або учасників семінару.

::::::::::::

::::::::::::::::::::::::::::::::::::::: objectives

- Опис формату даних JSON
- Розуміння, де JSON зазвичай використовується
- Оцініть деякі переваги використання JSON перед табличними даними
- Оцініть деякі недоліки обробки документів JSON
- Використовуйте пакет JSONLite для читання файлу JSON
- Відображення форматованого JSON як датафрейм
- Вибір та відображення вкладених полів датафрейму з документа JSON
- Запис табличних даних з вибраних елементів з документа JSON у файл csv

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- Що таке формат JSON?
- Як я можу перетворити JSON в датафрейм в R?
- Як я можу перетворити масив записів JSON у таблицю?

::::::::::::::::::::::::::::::::::::::::::::::::::

## Формат даних JSON

The JSON data format was designed as a way of allowing different machines or processes within machines to communicate with each other by sending messages constructed in a well defined format. JSON тепер є кращим форматом даних, який використовується API (Application Programming Interfaces).

Формат JSON, хоча і дещо детальний, не тільки читається людиною, але його також можна дуже легко зіставити на датафрейм R.

Ми збираємось прочитати файл даних, відформатований як JSON, перетворити його в датафрейм у R, а потім вибірково створити файл csv з витягнутих даних.

Файл JSON, який ми збираємось використовувати, - це файл [Safi.json] (data/SAFI.json). Це вихідний файл з електронної системи обстеження під назвою ODK. JSON представляє відповіді на серію запитань опитування. Самі питання замінені унікальними ключами, значення - відповіді.

Оскільки детальні опитування за своєю природою є вкладеними структурами, що дає змогу фіксувати різні рівні деталізації або вибірково ставити певні запитання залежно від відповіді на попереднє, структура відповідей в опитуванні може бути не лише складною й заплутаною, а й легко відрізнятися між різними респондентами.

### Переваги JSON

- Дуже популярний формат даних для API (наприклад, результати пошуку в Інтернеті)
- Читабельний
- Кожен запис (або документ, як їх називають) є самостійним. Еквівалент назви стовпця та значень стовпців є у кожному записі.
- Не всі документи повинні мати однакову структуру в одному файлі
- Структури документів можуть бути складними і вкладеними

### Недоліки JSON

- Це більш детальні дані, ніж еквівалентні дані у форматі csv
- Може бути складніше для обробки і відображення даних, форматованих в csv

## Використовуйте пакет JSONLite для читання файлу JSON

```{r, message=FALSE}
library(jsonlite)
```

Як і при читанні в CSV, у вас є кілька варіантів доступу до файлу JSON.

Ви можете прочитати файл JSON безпосередньо в R за допомогою `read_json () `або порівнянної функції `fromJSON ()`, хоча це не завантажує файл.

```{r eval=FALSE}
json_data <- read_json( "https://raw.githubusercontent.com/datacarpentry/r-socialsci/main/episodes/data/SAFI.json")
```

Щоб завантажити файл, ви можете скопіювати та вставити вміст файлу на
[GitHub] (https://github.com/datacarpentry/r-socialsci/blob/main/episodes/data/SAFI.json),, створивши файл `safi.json` у вашому каталозі `data`, або ви можете завантажити файл за допомогою R.

```{r download-data, eval=FALSE}
download.file(  "https://raw.githubusercontent.com/datacarpentry/r-socialsci/main/episodes/data/SAFI.json", "data/SAFI_clean.csv", mode = "wb"")
```

Після завантаження даних ви можете прочитати їх у R за допомогою `read_json () `:

```{r eval=FALSE}
json_data <- read_json("data/SAFI.json")
```

Ми бачимо, що в нашому середовищі з'явився новий об'єкт під назвою json\ _data. It is described as a Large list (131 elements). In this current form, our data is messy. You can have a glimpse of it with the `head()` or `view()` functions. It will look not much more structured than if you were to open the JSON file with a text editor.

This is because, by default, the `read_json()` function's parameter `simplifyVector`, which specifies whether or not to simplify vectors is set to FALSE. This means that the default setting does not simplify nested lists into vectors and data frames. However, we can set this to TRUE, and our data will be read directly as a dataframe:

```{r}
json_data <- read_json("data/SAFI.json", simplifyVector = TRUE)
```

Now we can see we have this json data in a dataframe format. For consistency with the rest of
the lesson, let's coerce it to be a tibble and use `glimpse` to take a peek
inside (these functions were loaded by `library(tidyverse)`):

```{r}
json_data <- json_data %>% as_tibble()
glimpse(json_data)
```

Looking good, but you might notice that actually we have a variable, _F\_liv_ that is a list of dataframes! It is very important to know what you are expecting from your data to be able to look for things like this. For example, if you are getting your JSON from an API, have a look at the API documentation, so you know what to look for.

Often when we have a very large number of columns, it can become difficult to determine all the variables which may require some special attention, like lists. Fortunately, we can use special verbs like `where` to quickly select all the list columns.

```{r}
json_data %>%
    select(where(is.list)) %>%
    glimpse()
```

So what can we do about _F\_liv_, the column of dataframes? Well first things first, we can access each one. For  example to access the dataframe in the first row, we can use the  bracket (`[`) subsetting. Here we use single bracket, but you could also use double bracket (`[[`). The `[[` form allows only a single element to be selected using integer or character indices, whereas `[` allows indexing by vectors.

```{r}
json_data$F_liv[1]
```

We can also choose to view the nested dataframes at all the rows of our main dataframe where a particular condition is met (for example where the value for the variable _C06\_rooms_ is equal to 4):

```{r}
json_data$F_liv[which(json_data$C06_rooms == 4)]
```

## Write the JSON file to csv

If we try to write our json\_data dataframe to a csv as we would usually in a regular dataframe, we won't get the desired result. Using the `write_csv` function from the `{readr}` package won't give you an error for list columns, but you'll only see missing (i.e. `NA`) values in these columns. Let's try it out to confirm:

```{r, eval=FALSE}
write_csv(json_data, "json_data_with_list_columns.csv")
read_csv("json_data_with_list_columns.csv")
```

To write out as a csv while maintaining the data within the list columns, we will need to "flatten" these columns. One way to do this is to convert these list columns into character types. (However, we don't want to change the data types for any of the other columns). Here's one way to do this using tidyverse. This command only applies the `as.character` command to those columns 'where' `is.list` is `TRUE`.

```{r}
flattened_json_data <- json_data %>% 
  mutate(across(where(is.list), as.character))
flattened_json_data
```

Now you can write this to a csv file:

```{r, eval=FALSE}
write_csv(flattened_json_data, "data_output/json_data_with_flattened_list_columns.csv")
```

Note: this means that when you read this csv back into R, the column of the nested dataframes will now be read in as a character vector. Converting it back to list to extract elements might be complicated, so it is probably better to keep storing these data in a JSON format if you will have to do this.

You can also write out the individual nested dataframes to a csv. For example:

```{r, eval=FALSE}
write_csv(json_data$F_liv[[1]], "data_output/F_liv_row1.csv")
```

:::::::::::::::::::::::::::::::::::::::: keypoints

- JSON is a popular data format for transferring data used by a great many Web based APIs
- The complex structure of a JSON document means that it cannot easily be 'flattened' into tabular data
- We can use R code to extract values of interest and place them in a csv file

::::::::::::::::::::::::::::::::::::::::::::::::::


